---
title: "The rational impression account of trust in science"
# If blank, the running header is the title in upper case.
shorttitle: ""
# Set names and affiliations.
# It is nice to specify everyone's orcid, if possible.
# There can be only one corresponding author, but declaring one is optional.
author:
  - name: Jan Pfänder
    corresponding: false
    orcid: 0009-0009-4389-2807
    affiliations:
      - id: id1  # Added an explicit ID for referencing
        name: ENS, EHESS, PSL University, CNRS, France
        department: Institut Jean Nicod, Département d’études cognitives

  - name: Hugo Mercier
    corresponding: true
    orcid: 0000-0002-0575-7913
    email: hugo.mercier@gmail.com
    affiliations: 
      - ref: id1  
blank-lines-above-author-note: 2
author-note:
  status-changes: 
    # Example: [Author name] is now at [affiliation].
    affiliation-change: ~
    # Example: [Author name] is deceased.
    deceased: ~
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    # Example: This study was registered at X (Identifier Y).
    study-registration: ~
    # Acknowledge and cite data/materials to be shared.
    data-sharing: ~
    # Example: This article is based on data published in [Reference].
    # Example: This article is based on the dissertation completed by [citation].  
    related-report: ~
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: The authors have no conflicts of interest to disclose.
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: ~
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
abstract: |
  Recognizing and addressing some of the pressing challenges we face as human society, including global health and climate change, requires trust in science. Philosophers of science have argued that people should trust science for its epistemic qualities–its capacity to produce accurate knowledge. Under this premise, the literature on public understanding of science has long sought to explain people’s trust in science by their knowledge of it–with sobering results: While people do tend to trust science, they do not tend to know much about it. If not grounded in knowledge, is public trust in science mostly irrational? Here, we argue that no, not necessarily. We propose an explanation of the foundations of trust in science at the micro-level: the ‘rational impression’ account of trust in science. The account builds on two basic mechanisms of information evaluation: First, if someone finds out something that is hard-to-know, we tend to be impressed by it, if we deem it true. This impression makes us infer that the person is competent, a crucial component of trustworthiness. However, it is hard or even impossible to recall exactly how we formed this impression. Second, if something is highly consensual, it is likely to be true. This is particularly relevant when people lack relevant background knowledge to evaluate claims for themselves, as is often the case in science. Our review underlines the critical role of education and science communication in fostering trust in science.
  
# Put as many keywords at you like, separated by commmas (e.g., [reliability, validity, generalizability])
keywords: [trust in science, science literacy, deficit model]
# If true, tables and figures are mingled with the text instead of listed at the end of the document.
impact-statement: ~
# If true, a word count will appear below the keywords (tables, figure captions, and references excluded in count.)
word-count: true
floatsintext: true
# Numbered lines (.pdf and .docx only)
numbered-lines: false
# File with references
bibliography: bibliography.bib
# Suppress title page
suppress-title-page: false
# Link citations to references
link-citations: true
# Masks references that appear in the masked-citations list
mask: false
masked-citations:
# If true, adds today's date below author affiliations. If text, can be any value.
# This is not standard APA format, but it is convenient.
# Works with docx, html, and typst. 
draft-date: true
# Language options. See https://quarto.org/docs/authoring/language.html
lang: en
language:
  citation-last-author-separator: "and"
  citation-masked-author: "Masked Citation"
  citation-masked-date: "n.d."
  citation-masked-title: "Masked Title"
  email: "Email"
  title-block-author-note: "Author Note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; [credit.niso.org](https://credit.niso.org)) as follows:"
  title-impact-statement: "Impact Statement"
  references-meta-analysis: "References marked with an asterisk indicate studies included in the meta-analysis."
format:
  apaquarto-docx: 
    toc: false
  apaquarto-html: 
    toc: true
  apaquarto-typst: 
    keep-typ: true
    list-of-figures: false
    list-of-tables: false
    toc: false
    papersize: "us-letter"
  apaquarto-pdf:
    # Can be jou (journal), man (manuscript), stu (student), or doc (document)
    documentmode: jou
    keep-tex: true

header-includes:  | # to prevent floats from moving past certain points (for the appendix)
  \usepackage{placeins} 
    
always_allow_html: true
---

```{r}
#| label: setup
#| include: false
#| message: false
library(tidyverse)
library(flextable)
library(tinytable)
library(kableExtra)

```

## Introduction

Addressing important societal challenges, from fighting climate change to managing pandemics, requires trust in science. Numerous studies have demonstrated that individuals with higher levels of trust in science are more likely to engage in pro-environmental behavior, support climate policies, and accept the scientific consensus on global warming [@colognaRoleTrustClimate2020]; @hornseyMetaanalysesDeterminantsOutcomes2016; @bogertEffectTrustScience2024\]. Trust in science has also been shown to be positively associated with willingness to get vaccinated, both in a Covid-19 context [@lindholtPublicAcceptanceCOVID192021] and beyond [@sturgisTrustScienceSocial2021]. During the Covid-19 pandemic, a panel study in 12 countries found that trust in scientists was the strongest predictor of whether people followed public health guidelines, such as mask-wearing or social distancing [@alganTrustScientistsTimes2021]. Similar results have been found by other studies, e.g. positive effects of trust in science on acceptance of social distancing in the US [@koetkeTrustScienceIncreases2021].

How much do people trust or do not trust science? And why? These questions are critical for a wide range of public policies that are informed by scientific evidence.

Here, we begin by showing that the literature on public understanding of science has long focused on why people do not trust science. The most popular explanation has been that people do not know enough about science. We refer to this as the knowledge account of trust in science.

We then lay out the puzzle of trust in science: Across the globe, most people do tend to trust most of science, despite not knowing much about it. This does not quite match the predictions of the knowledge account.

If not grounded in knowledge, is public trust in science mostly irrational? We argue that no, not necessarily. We develop an alternative explanation: the 'rational impression' account of trust in science. According to this account, people do not need a profound understanding or detailed knowledge of science, to perceive it as trustworthy. Instead, science impresses people, by appealing to basic cognitive mechanisms of information evaluation.

The rational impression account is not built on the premise that all is well regarding public trust in science. Although most people tend to trust science, this trust is far from being at ceiling. Besides, distrusting minorities can cause serious societal harm, especially when in positions of political power. Nor does the rational impression account claim to explain why people distrust science.

## The puzzle of why people trust science

A widely agreed-upon definition of trust is the willingness to be vulnerable to another party–whether an individual, a group, or an institution [@mayerIntegrativeModelOrganizational1995a; @rousseauIntroductionSpecialTopic1998]. Building on this idea, trust in science has been defined as “one’s willingness to rely on science and scientists (as representatives of the system) despite having a bounded understanding of science” [@wintterlinPredictingPublicTrust2022, p. 2].

This definition implies that trust in science goes beyond knowledge of science. Yet, the idea that knowledge of science is the primary cause of trust in–and more generally positive attitudes towards–science, has dominated research on public understanding of science [@bauerWhatCanWe2007]. This idea is widely known under the term "deficit model", because much of the literature attested to the public "depressingly low levels of scientific knowledge" that were assumed to be the principal cause of negative attitudes towards science [@sturgisScienceSocietyReEvaluating2004, p. 56]. Here, we will refer to the idea that knowledge of science causes trust in science with a more neutral term: knowledge account of trust in science.

In this section, we will showcase that the knowledge account does not align with the data: People trust science, to a large extent, despite not knowing much about it. Although many have criticized the knowledge account, there are currently no alternative accounts that seek to explain the elevated levels of public trust in science.

### People tend to trust science

Across the globe, most people do trust science, at least to some extent. A recent study in 68 countries found that, across the globe, trust in scientists was "moderately high" (mean = 3.62; sd= 0.70; Scale: 1 = very low, 2 = somewhat low, 3 = neither high nor low, 4 = somewhat high, 5 = very high), with not a single country below midpoint trust [@colognaTrustScientistsTheir2025]. Long-term global data on trust in science across time is sparse. Yet, the available data suggests, if anything, a recent increase of trust in science: In 2018, the Wellcome Global Monitor (WGM) surveyed of over 140000 people in over 140 countries on trust in science [@wellcomeglobalmonitorWellcomeGlobalMonitor2018]. In 2020, during the first year of the Covid pandemic and before vaccines were widely available, a follow-up survey was run in 113 countries, involving 119000 participants [@wellcomeglobalmonitorWellcomeGlobalMonitor2020]. Between these two surveys, on average, trust in science has risen [@wellcomeglobalmonitorPublicTrustScientists2021]: In 2020, 41% (32% in 2018) of respondents said they trust science a lot, 39% (45% in 2018) said they trust science to some extent, 13% (also 13% in 2018) said they trust science "not much or not at all", and 7% (10% in 2018) answered "don't know". In the US, where long term data is available from the US General Social Survey (GSS), this public trust appears to be both remarkably stable and elevated relative to other institutions [@funkPublicConfidenceScientists2020; @funkScienceScientistsHeld2020; @smithTrendsPublicAttitudes2013]: From the early 1970s to 2022, the currently latest year available in the GSS, on average 40% of Americans say they have a great deal of confidence in the scientific community. This is the second highest score (just behind the military) among 13 institutions listed in the GSS, including, e.g., the government, press, organized religion or medicine. Note, however, that the most recent polls suggest a drop in trust in science in the US [@lupiaTrendsUSPublic2024].

### People do not know much about science

The knowledge account of trust in science posits that knowledge is the main cause of trust in science. Accordingly, the elevated levels of trust in science should be matched by elevated levels of science knowledge. But this is not the case.

Early attempts of measuring science knowledge have developed what is known as the "Oxford scale" (@tbl-oxford[^1]) to measure science knowledge–a set of specific true/false or multiple-choice questions about basic science facts.[^2] As the name suggests, this measure was developed by scholars associated with the University of Oxford in the late 1980s, in collaboration with Jon D. Miller, a pioneer in research on science literacy in the US [@durantPublicUnderstandingScience1989]. The Oxford scale remains popular: different versions of it have been used in large-scale survey projects, including several Eurobarometer surveys, as well as in the National Science Foundation's (NSF) "Science and Engineering Indicators" surveys in the US [@nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016]. Survey results from the Oxford scale have been taken to showcase a science knowledge deficit among the public. For example, from the first surveys in which they were used, @durantPublicUnderstandingScience1989 (p.11) report that only "34% of Britons and 46% of Americans appeared to know that the Earth goes round the Sun once a year, and just 28% of Britons and 25% of Americans knew that antibiotics are ineffective against viruses". According to the @nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016, performance on the Oxford scale items in the US has been "fairly stable across 2 decades" [@nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016, p. 51][^3].

[^1]: The original scale proposed by @durantPublicUnderstandingScience1989 comprised 20-items. Several reduced versions of this have been used in different survey projects and studies since then. For example, @millerMeasurementCivicScientific1998a reports that a Eurobarometer in 1992 included only 9, and a Science and Engineering Indicator survey in 1995 only 10 of these items. @gauchatCulturalAuthorityScience2011 reports relying on a 14-item scale based on several Science and Engineering Indicators surveys included in the GSS (but does not provide an overview of the included items).

[^2]: The meta-analysis included two types of studies: one that measured general scientific knowledge, and one that measured knowledge specific to biology and genetics

[^3]: Note that for this trend scale, the National Science Board who publishes results from the Science and Engineering Indicators selected a subset of 9 of the items shown in [@tbl-oxford; for details, see @nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016].

::: {#tbl-oxford}
```{r}
# Create the data frame
knowledge_items <- data.frame(
  `Question` = c(
    "The center of the Earth is very hot. (True)",
    "The continents on which we live have been moving their locations for millions of years and will continue to move in the future. (True)",
    "Does the Earth go around the Sun, or does the Sun go around the Earth? (Earth around Sun)",
    "How long does it take for the Earth to go around the Sun? (One year)\\*",
    "All radioactivity is man-made. (False)",
    "It is the father’s gene that decides whether the baby is a boy or a girl. (True)",
    "Antibiotics kill viruses as well as bacteria. (False)",
    "Electrons are smaller than atoms. (True)",
    "Lasers work by focusing sound waves. (False)",
    "Human beings, as we know them today, developed from earlier species of animals. (True)",
    "The universe began with a huge explosion. (True)"
  )
) |> 
  mutate(number = 1:11) |> 
  select(number, Question)

# Render the table
kable(
  knowledge_items,
  col.names = c("", ""),
  booktabs = TRUE, 
      longtable = TRUE,
      #caption = "Science knowledge items",
      full_width = TRUE
) %>%
  kable_styling() %>%
    kableExtra::column_spec(1, width = "2em") %>%
    kableExtra::column_spec(2, width = "40em") %>%
  footnote(
    general = "*Only asked if previous question was answered correctly.",
    general_title = ""
  )
```

An 11-item version of the Oxford-scale, as reported in a comprehensive review of the literature on scientific literacy [@nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016].
:::

The Oxford scale has received various forms of criticism, ranging from minor concerns for being--ironically, in light of the bad performance--too easy to answer [@kahanClimateScienceCommunicationMeasurement2015], to the fundamental critique that it only captures factual recall [@pardoCognitiveDimensionPublic2004; @bauerWhatCanWe2007]. What should actually matter for trust, according to this fundamental critique, is a different kind of knowledge, namely an institutional and methodological understanding of how science works.

The literature on science literacy goes well beyond Oxford-scale type, factual measures of trust in science. Even the creators of the Oxford scale were, to some extent, aware of its limitations [@nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016]. They never intended the Oxford scale items to serve as a comprehensive measure of science literacy. For example, @durantPublicUnderstandingScience1989 also developed a scale of "understanding of processes of scientific inquiry"--several multiple-choice questions about the scientific method and basic concepts of probability. @millerPublicUnderstandingAttitudes2004 viewed a scientifically literate citizen as someone who has both a “(1) a basic vocabulary of scientific terms and constructs; and (2) a general understanding of the nature of scientific inquiry.” His measure of science literacy included open-ended questions, for example on what people understand as the meaning of scientific study [@millerMeasurementCivicScientific1998a].

However, these measures have hardly drawn a more positive image of the public's knowledge of science: Using an index of various understanding questions, @millerPublicUnderstandingAttitudes2004 (p. 288) concluded that "approximately 10 percent of US adults qualified as civic scientifically literate in the late 1980s and early 1990s, but this proportion increased to 17 percent in 1999". Miller described that, according to his measure, someone qualifies as scientifically literate if they possess "the level of skill required to read most of the articles in the Tuesday science section of The New York Times, watch and understand most episodes of Nova, or read and understand many of the popular science books sold in bookstores today" [@millerPublicUnderstandingAttitudes2004, p. 288].

More recent data suggest that science literacy in the US may have improved slightly since Miller's assessment during the early 2000s, but it still remains rather low. Based on results from the 2018 US Science & Engineering Indicators, @scheufeleScienceAudiencesMisinformation2019 (p. 7663) report that "one in three Americans (36%) misunderstood the concept of probability; half of the population (49%) was unable to provide a correct description of a scientific experiment; and three in four (77%) were unable to describe the idea of a scientific study."

The 2024 US Science & Engineering Indicators, based on data from the Pew Research Center’s American Trends Panel (ATP) from 2020, report that "60% of U.S. adults could correctly note that a control group can be useful in making sense of study results" and that "only half of U.S. adults (50%) could correctly identify a scientific hypothesis" [@nationalscienceboardnationalsciencefoundationScienceTechnologyPublic2024, p.24].

### A weak correlation between knowledge and trust

So far, we have contrasted the elevated levels of trust with the low levels of science knowledge. This is only indirect evidence for a weak association between the two. More direct evidence on this association comes with two major limitations: First, direct evidence is based mostly on narrow, factual knowledge as measured by the Oxford scale, and not on a broader, institutional understanding of science. Second, the relevant studies have typically assessed attitudes towards science more broadly, rather than trust in science in particular. In a seminal meta-analysis @allumScienceKnowledgeAttitudes2008 found that Oxford scale type science knowledge was only weakly associated with attitudes towards science. Moreover, this weak association only held for general attitudes–for attitudes towards specific contentious science topics (e.g., climate change), the study did not find an association.

Another limitation of the evidence of a weak association between trust in science and science knowledge presented so far is that it is–to a large extent–based on data from the US and, to some extent, Europe. However, a recent global study points towards a similar conclusion: @colognaTrustScientistsTheir2025 tested the relationship between national science literacy scores, based on the Program for International Student Assessment (PISA), and national average trust in scientists for the 68 countries included in their study. They found no statistically significant association.

Why, if not for knowing about science, do people tend to trust science then? As we will argue in the next section, this puzzle remains largely unaddressed.

### Other accounts of trust in science

The knowledge account of trust in science has been widely criticized for idealizing of science and viewing the public as deficient: for implying that "to know science is to love it" [@bauerWhatCanWe2007] and for portraying science knowledge as "superior to whatever 'nonscientific' or 'local' knowledge the public may (also) possess" [@gauchatCulturalAuthorityScience2011]. The literature has long moved beyond the idea of science knowledge as the principle driver of trust in science. However, the focus on explaining a lack of trust, rather than trust, persists.

Researchers have increasingly studied how people's values, world views, and identities shape their attitudes towards science [@hornseyAttitudeRootsJiu2017a; @lewandowskyWorldviewmotivatedRejectionScience2021]. The psychological literature has focused on explaining negative attitudes towards science with motivated reasoning–selecting and interpreting information to match one's existing beliefs or behaviors [@lewandowskyMotivatedRejectionScience2016; @hornseyWhyFactsAre2020; @lewandowskyRoleConspiracistIdeation2013]. This research mostly suggests that certain psychological traits, such as a social dominance orientation, or a tendency of engaging in conspiracy thinking, lead people to reject science. Arguments on a general conspiratory thinking style as one of the root causes of science rejection shift the debate, to some extent, from a knowledge deficit to a broader reasoning deficit [@hornseyAttitudeRootsJiu2017a; @rutjensConspiracyBeliefsScience2022][^4].

[^4]: Note that in these studies, conspiracy thinking is conceived of as a general psychological trait, i.e. a general a way of thinking. For a distinction between general conspiracist worldviews and conspiracy beliefs about science specifically, see @rutjensConspiracyBeliefsScience2022.

Motivated reasoning accounts have been popular in light of accumulating evidence in the US for a widening partisan gap regarding trust in science, with Republicans trusting less and Democrats trusting more [@gauchatPoliticizationSciencePublic2012; @krauseTrendsAmericansTrust2019a; @leePartyPolarizationTrust2021]. Contrary to what the deficit model would suggest, some influential work on motivated reasoning has shown that partisan-driven rejection of science does not appear to be the result of a lack of cognitive sophistication: @kahanPolarizingImpactScience2012 have shown that greater science literacy was associated with more polarized beliefs on climate change. @drummondIndividualsGreaterScience2017 have extended these findings to other controversial science topics, namely stem cell research and evolution: they show that both greater science literacy and education are associated with more polarized beliefs on these topics. However, this phenomenon that "people with high reasoning capacity will use that capacity selectively to process information in a manner that protects their own valued beliefs" [@perssonPreregisteredReplicationMotivated2021, p.1], known under the term 'motivated numeracy', has largely failed to replicate [@perssonPreregisteredReplicationMotivated2021; @hutmacherMotivatedReasoningClimate2024; @stagnaroNoAssociationNumerical2023a]. Other studies, focusing on the case of climate change, have argued that partisan divides might simply be the result of bayesian information updating, rather than motivated reasoning [@bayesMotivatedReasoningClimate2021; @druckmanEvidenceMotivatedReasoning2019].

Since the Covid-19 pandemic, the role of misinformation in fostering distrust in science has been increasingly studied [@nationalacademiesofsciencesUnderstandingAddressingMisinformation2024; @scheufeleScienceAudiencesMisinformation2019; @druckmanThreatsSciencePoliticization2022]. For this literature, by contrast with the deficit model, the problem of trust in science is less a lack of information, and more the abundance of harmful information.

The premise of the deficit model was that people would trust science because of their knowledge and understanding of science. Perhaps in response to this, recent literature in science communication has shifted the focus on source-based evaluations of trust: from science knowledge to perceptions of scientists. This work has established that people evaluate scientists along different dimensions [@intemannScienceCommunicationPublic2023], including competence, but also also integrity, benevolence or openness [@hendriksMeasuringLaypeoplesTrust2015; @besleyReassessingVariablesUsed2021a]. This literature suggests that, for enhancing trust in science, the latter, warmth-related (i.e. other than competence) dimensions could be particularly relevant [@fiskeGainingTrustWell2014]. The idea is that people already perceive scientists as very competent, but not as very warm, thus offering a greater margin for improvement.

Another explanation for distrust towards science is the *alienation model* [@gauchatCulturalAuthorityScience2011]. According to this model, the "public disassociation with science is a symptom of a general disenchantment with late modernity, mainly, the limitations associated with codified expertise, rational bureaucracy, and institutional authority” [@gauchatCulturalAuthorityScience2011, p.2]. This explanation builds on the work of social theorists [@habermasJurgenHabermasSociety1989; @beckRiskSocietyNew1992; @giddensModernitySelfidentitySelf1991; see @gauchatCulturalAuthorityScience2011 for an overview] who suggested that a modern, complex world increasingly requires expertise, and thus shapes institutions of knowledge elites. People who are not part of these institutions experience a lack of agency, resulting in a feeling of alienation. If the alienation model might account for some of the distrust towards science, its goal is not to explain why most people still trust science.

On the whole, the literature on public trust in science focuses not on explaining why people do trust science, but on why they do not, or not enough. As reviews of science-society research have noted, the literature operates in "deficit" paradigms [@bauerWhatCanWe2007; @scheufeleThirtyYearsScience2022] which tend to overlook the elevated levels of trust.

## The rational impression account of trust in science

The previous section has established that public trust in science is relatively high, but that knowledge and understanding of science do not seem to be strong determinants of this trust.

In this section, we try to make sense of this by taking a cognitive perspective. We develop a 'rational impression' account of trust in science, according to which people trust science because they have been impressed by it. This impression of trust persists even after knowledge of the specific content has vanished. The account builds on two basic mechanisms of information evaluation: First, if something is highly consensual, it is likely to be true. This is particularly relevant when people lack relevant background knowledge to evaluate claims for themselves, as is often the case in science. Second, if someone finds out something that is hard-to-know, we tend to be impressed by it, if we deem it true. This impression makes us infer that the person is competent, a crucial component of trustworthiness. However, it is hard or even impossible to recall exactly how we formed this impression.

#### People infer accuracy from consensus

In order to make the best of communicated information, animals need to be able to evaluate it, i.e. being able to distinguish inaccurate and harmful from accurate and beneficial information [@maynard-smithAnimalSignals2003]. It has been argued that humans have evolved a suite of cognitive mechanisms to serve this function [@sperberEpistemicVigilance2010; @mercierNotBornYesterday2020]. In particular, we rely on cues of an informant's trustworthiness, and check the plausibility of an information against our background knowledge.

In the case of science, reliable cues and background information are scarce: people generally have little first-hand information to evaluate individual scientists' trustworthiness, because they don't know scientists personally. People also largely lack relevant background knowledge to evaluate the plausibility of scientific findings. Sometimes, to a certain extent, people might be able to judge the accuracy of scientific findings for themselves, for example when they are exposed to accessible and convincing explanations in school [@readExplanatoryCoherenceSocial1993; @lombrozoSimplicityProbabilityCausal2007; for a review, see @lombrozoStructureFunctionExplanations2006]. But for most scientific research, people cannot possibly evaluate the quality of the information for themselves, let alone make their own observations (e.g. quantum mechanics, genes).

An additional way to evaluate whether something is true or not is to aggregate opinions. It has been shown that, when no better information is available, people rely on majority heuristics: the more others agree on something, the more likely we are to believe them to be right [@mercierMajorityRulesHow2019]. Literature on the wisdom of crowds has shown that this inference–from convergence of opinions to accuracy–is often appropriate [see e.g., @hastieRobustBeautyMajority2005]. In non-science related contexts, it has been shown that people go even further and infer that others are more competent, the more they agree with each other [@pfanderHowWiseCrowd2025]. In these experiments, participants were deprived of relevant background knowledge and were told that the others were answering independently of each other.

More indirect evidence suggests that people make this inference also in the context of science: @pfanderFrenchTrustMore2025 showed that in France, people trust scientists more when they work in disciplines that people perceive as more consensual. Yet more suggestive evidence comes from a popular psychological model, the "gateway model". The model suggests that informing people about the scientific consensus on specific issues acts as a gateway to change their beliefs on these issues [@vanderlindenGatewayBeliefModel2021]. Studies have demonstrated the effectiveness of consensus messaging in changing people's beliefs on contentious science topics such as climate change [@veckalov27countryTestCommunicating2024] or vaccination [@salmonVaccineHesitancyCauses2015; for an overview of results on vaccination, climate change, and genetically modified food, see @vanstekelenburgScientificConsensusCommunicationContested2022]. This evidence is only suggestive, however, because the fact that consensus changes people's beliefs or attitudes does not necessarily require enhanced trust. An alternative explanation, for example, is normative conformity-that is, when people follow the majority because of social pressure rather than a belief that the majority is correct [@mercierMajorityRulesHow2019].[^5]

[^5]: However, an accuracy inference seems to be the more plausible mechanism here: Studies on consensus messaging do not seem to be settings of high social pressure that we might expect to produce instances of normative conformity, compared to, for instance, the famous Asch experiments [@aschStudiesIndependenceConformity1956].

In the absence of other reliable cues and background knowledge, inferences from consensus are likely to be given considerable weight in a cognitive system of epistemic vigilance. In the case of science, this weight should play in favor of science's perceived trustworthiness: It has been argued that, by contrast with other intellectual enterprises, consensus is the defining trait of science [@collinsSociologyPhilosophiesGlobal2002]. Not only do scientists agree on things, but they agree on impressive things–things that would be impossible for any individual to ever uncover for themselves.

#### People trust but forget impressive science

People use the degree of consensus as a cue to infer whether an information is accurate, and, to some extent, whether the informants are competent. However, this latter inference on the informants' competence should strongly depend on how impressive the information is.

For an information to be impressive, at least two criteria should be met: (i) there is reason to believe it is true, and (ii) it is perceived as hard to uncover. In the case of science, perceived consensus might be the main relevant cue for (i), and (ii) can be mostly taken for granted. For example, most people would probably only be mildly impressed by someone telling them that a given tree has exactly 110201 leaves. Even though obtaining this information implies an exhausting counting effort, everyone in principle knows how to do it. By contrast, finding out that it takes light [approximately 100,000 years to travel from one end of the Milky Way to the other](https://imagine.gsfc.nasa.gov/features/cosmic/milkyway_info.html) is probably impressive to most people, as they would not know how such a distance can be measured.

Outside the realm of science, with trivia questions, it has been shown that people have accurate perceptions of whether something is hard to know or not, and that they use this information to infer someone's competence [@dubourgUsingNestedStructure2025]: knowing a rare piece of information indicates a high likelihood of knowing more information in the same domain.

In the case of science, @pfanderTrustingForgettingImpressive2025 showed that reading about impressive scientific findings increased participants' perceptions of both the scientists' competence and the trustworthiness of their discipline. At the same time, participants forgot almost immediately about the specific content that generated these impressions.

This forgetting could explain the low levels of science knowledge. We commonly form impressions of the people around us while forgetting the details of how we formed these impressions: If a colleague fixes our computer, we might forget exactly how they fixed it, yet remember that they are good at fixing computers. As an extreme example, patients with severe amnesia can continue to experience emotions linked to events they could not recall [@feinsteinSustainedExperienceEmotion2010]. In the context of science, @liquinMotivatedLearnAccount2022 have shown that while people find some science-related explanations more satisfying than others, this did not predict how well they could recall the explanations shortly after, suggesting that impressions and knowledge formation can be quite detached.

### Additional predictions of the rational impressions account

The rational impressions account makes several additional predictions.

First, competence should be the main dimension of scientists' trustworthiness. In social psychology, a popular model suggests that people evaluate others along two fundamental dimensions: competence and warmth [@cuddyWarmthCompetenceUniversal2008]. Similarly, for trust in scientists, researchers have distinguished between an epistemological and an ethical dimension [@wilholtEpistemicTrustScience2013; @intemannScienceCommunicationPublic2023]. Sometimes, researchers make more fine-grained distinctions: For example, @hendriksMeasuringLaypeoplesTrust2015 have argued for three dimensions: expertise/competence, integrity, and benevolence. @besleyReassessingVariablesUsed2021a has suggested openness as an additional fourth dimension. Across these dimensions, anything but competence should be very hard to evaluate for people: there are relatively very few scientists in the world, and most people probably do not know any personally. The only other way they could judge scientists' character is through media coverage. But news on science–by contrast, for example, with news on politicians–mostly tend to concern the science, not the scientists. Besides, people consume very little news in general [@newmanDigitalNewsReport2023]. Competence, however, can be judged based on the mechanisms of the rational impression account. Accordingly, scientists should score higher in competence evaluations than in other dimensions of trustworthiness. In line with this prediction, it has been shown that people perceive scientists as very competent, but not so much as warm [@fiskeGainingTrustWell2014]. A recent Pew survey found that 89% of Americans viewed research scientists as intelligent, but only 65% viewed them as honest, and only 45% described research scientists as good communicators [@kennedyPublicTrustScientists2024; see also @fiskeGainingTrustWell2014]. Beyond the US, a recent study confirmed this tendency on a global scale [@colognaTrustScientistsTheir2025]: People perceived scientists as highly competent, with 78% tending to believe that scientists are qualified to conduct high-impact research. By contrast, people held scientists in lower esteem with regards to their integrity and benevolence: Only 57% of people tended to believe that most scientists are honest, and only 56% tended to believe that most scientists are concerned about people’s well-being.

Second, education, and more precisely science education, should be the main correlate of trust in science. Since most people consume very little news [@newmanDigitalNewsReport2023], the bulk of exposure to science can be assumed to happen during education. Education, and in particular science education, has been consistently identified as one of the strongest correlates of trust in science [@noyScienceGoodEffects2019; @wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020; but see @colognaTrustScientistsTheir2025 who only find a small positive relationship between tertiary education and trust in science]. This is compatible with the fact that people, even those who received a science education, do not know much about science: if we assume that education has some causal effect on trust in science, this effect does not need to be driven by a pure transmission of knowledge and understanding [for a similar argument, see @bakEducationPublicAttitudes2001]. The candidate mechanism proposed by the rational impression account is exposure to impressive scientific content. Students might not understand much of it, and potentially recall even less later on; but they might have been impressed by it, to the point that they come to perceive scientists as competent, and thus, everything else equal, as trustworthy. This impression might persist even when specific knowledge vanishes. In line with this, @mottaEnduringEffectScientific2018 found that, in the US, the more children were interested in science at age 12–14 years, the more they tended to trust in climate scientists in adulthood (mid thirties), irrespective of their political ideology.

Third, people with a basic science education should trust essentially all of basic science. These people should have had the opportunity to form impressions of trustworthiness of science. This should have built a solid baseline of trust in science. People might deviate from this default and distrust science on certain specific science topics for other reasons, but they should trust most of science. This is in line with the finding that in the US, almost everyone--even people who say they don't trust science in general or who hold specific beliefs blatantly violating scientific knowledge (e.g. that the earth is flat)--trusts almost all of basic science knowledge (e.g. that electrons are smaller than atoms) [@pfanderQuasiuniversalAcceptanceBasic2025].

## Discussion

It has long been a puzzle to the deficit model–which suggests that trust in science is primarily driven by science knowledge–that knowledge of science is at best weakly associated with science attitudes [@allumScienceKnowledgeAttitudes2008; @nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016]. The rational impression account can make sense of this: it lays out how trusting science without recalling specific knowledge can be the result of a sound inference process, rooted in basic cognitive mechanisms of information evaluation.

The account is compatible with the finding that education, and in particular science education, has been repeatedly identified as one of the strongest correlates of trust in science [@bakEducationPublicAttitudes2001; @noyScienceGoodEffects2019; @wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020; but see @colognaTrustScientistsTheir2025]. By contrast with the deficit model, it suggests that the main causal role of education for public trust in science is not transmission of knowledge and understanding, but impression generation.

The rational impression account aligns with recent normative accounts of what makes science trustworthy. Instead of particular institutional features-certain methods, norms, or processes–these accounts argue that the trustworthiness of science lies in its diversity: @cartwrightTangleScienceReliability2022 make the case that scientific knowledge emerges from a "tangle" of results, relying on diverse research methods. @oreskesWhyTrustScience2019 makes a similar case: She argues that scientific practice takes place in different scientific communities who rely on a variety of different research methods. Through some shared practices, in particular peer-review, these communities engage in critical dialogue. What makes scientific knowledge trustworthy, according to Oreskes, is when from this diversity of actors and methods, a consensus emerges. According to this view, to infer trustworthiness, people should have a representation of the diversity of science. The rational impression account is, in a way, less strict: it does not require a representation of diversity. It does require, however, that people have a representation of science as an institution of independent thinkers.

The rational impressions account faces several limitations. First, it proposes a possible micro-level model of trust in science and should be seen as complementing, not competing with, macro-level processes that shape public trust in science. The rational impression account fits with a sociological literature investigating how "individual cognition and practice establish and maintain institutional fields and status hierarchies, especially in the face of imperfect knowledge" [@gauchatCulturalCognitiveMappingScientific2018, p.569]. However, sociological macro-level accounts have described how trust in science is entangled with broader cultural and political dynamics. These accounts, like the individual-level accounts reviewed above, tend to focus on explaining distrust in science. For example, @gauchatCulturalAuthorityScience2011 describes the 'alienation model', according to which the "public disassociation with science is a symptom of a general disenchantment with late modernity, mainly, the limitations associated with codified expertise, rational bureaucracy, and institutional authority” [@gauchatCulturalAuthorityScience2011, p.2]. This explanation builds on the work of social theorists [@habermasJurgenHabermasSociety1989; @beckRiskSocietyNew1992; @giddensModernitySelfidentitySelf1991; see @gauchatCulturalAuthorityScience2011 for an overview] who suggested that a modern, complex world increasingly requires expertise, and thus shapes institutions of knowledge elites. People who are not part of these institutions experience a lack of agency, resulting in a feeling of alienation. Similarly, @gauchatLegitimacyScience2023 argues that politicization of science in the US needs to be seen in its broader cultural context. Precisely, according to Gauchat, science has enabled the authority of the modern regulatory state. Consequently, conservative distrust of science reflects deeper structural tensions with the institutions and rational–legal authority of modern governance. At the micro-level, this is consistent with research showing that right-wing authoritarian ideology is associated with distrust towards science and scientists [@kerrRightwingAuthoritarianismSocial2021].

A second limitation of the rational impression account is that it assumes people have a representation of science as consensual. However, in practice–with perhaps some exceptions, such as during the Covid-19 pandemic–most people do not literally compare the opinions of different scientists for themselves and come to the conclusion that something is largely consensual. Where, then, could the representation of consensus possibly emerge? A plausible explanation, we believe, is that education fosters a representation of consensus: During education, in particular during early education, knowledge is typically presented as simply the result of science–a seemingly unanimous enterprise that produces knowledge. School books hardly teach about historical science controversies, suggest uncertainty around scientific findings, or cover cutting-edge research where disagreements are the norm. This could induce a default consensus assumption in people's perceptions of science. However, this argument is of course only speculative.

Third, the rational impression account cannot explain, for example, why people with no education, and thus presumably very little exposure to science, have some trust in science [@wellcomeglobalmonitorWellcomeGlobalMonitor2018]. A possible explanation could be two step effects, via some educated people who trust science whom they trust.

Fourth, conversely, the account also cannot explain why, in a context of the global north, where essentially everyone has been exposed to science through a basic science education, some people do not trust some aspects of science, or say they don't trust science in general [even if that is not really true, see @pfanderQuasiuniversalAcceptanceBasic2025]. Suggestions have already been made for a number of issues such as vaccination ([Miton and Mercier 2015](https://janpfander.github.io/phd_thesis/references.html#ref-mitonCognitiveObstaclesProVaccination2015)), GMOs ([Blancke et al. 2015](https://janpfander.github.io/phd_thesis/references.html#ref-blanckeFatalAttractionIntuitive2015)), or nuclear energy ([Hacquin et al. 2021](https://janpfander.github.io/phd_thesis/references.html#ref-hacquinDisgustSensitivityPublic2021)). However, research is still needed to better understand what motivates these rejections (see e.g., [Hornsey 2020](https://janpfander.github.io/phd_thesis/references.html#ref-hornseyWhyFactsAre2020)).

Beyond these theoretical and empirical limitations, the rational impression account is limited in its implications. First, we do not believe that flooding people with impressive consensual science knowledge is the key to overcoming all distrust of science. In the context of trust in political institutions, it has been argued that trust and distrust are not necessarily symmetrical: what causes the former might not help alleviate the latter [@bertsouRethinkingPoliticalDistrust2019]. We believe this is at least to some degree true for science, too. For example, consensus messaging has been shown to help convince people to trust science on particular issues, such as climate change or vaccines, but it is less clear whether it worked by fostering perceptions of trustworthiness. It could be the case that the people convinced by consensus messages already trusted science, but have not held strong opinions on the specific matter. This is not implausible, since it has been shown that on most matters, large segment of the public do not have opinions [@bourdieuPublicOpinionDoes1979; @zallerNatureOriginsMass1992]. For people who do not only lack trust, but who actively distrust, motivated reasoning accounts are likely better suited as a theoretical framework. Addressing relevant underlying motivations directly might be more fruitful to mitigate distrust in science than exposing people to consensual science more generally.

Second, and related, just because we propose an account by which trust in science can be the result of a rational cognitive process, this does not imply that, conversely, all distrust in science is irrational. Some groups of people do in fact have good reasons not to trust science. For example, some science has historically contributed to fostering racism [see e.g. @fuentesSystemicRacismScience2023; @noblesScienceMustOvercome2022], via instances such as the tragically famous Tuskegee syphilis study [@brandtRacismResearchCase1978; @scharffMoreTuskegeeUnderstanding2010].

Third, we do not think that science communication should stress consensus at all costs. In the rational impression account, consensus plays a central role for generating trust. However, this should not incentivize science communicators to neglect transparency about uncertainty. Acknowledging uncertainty in science communication has been argued to be crucial for fostering long term trust in science [@druckmanCommunicatingPolicyRelevantScience2015]. For example, in the context of Covid-19 vaccines, @petersenTransparentCommunicationNegative2021 have shown that communicating uncertainty is crucial for building long term trust in health authorities.

Fourth, science communication should not aim for impressiveness at all costs either. Research has shown that intellectual humility can increase trust in scientists [@koetkeEffectSeeingScientists2024]. Trying to oversell scientific results might therefore backfire. People appear to value transparency via open data practices in science [@songTrustingShouldersOpen2022], and trust science that replicates more [@hendriksReplicationCrisisTrust2020]. We should therefore expect that simply doing better, more transparent science, and being humble about it, is likely to be the most effective strategy to impress the public and elicit perceptions of trustworthiness.

Fifth, educators should not stop aiming at fostering a proper understanding of science. Most students might not understand all of the content, or recall much specific knowledge later on. However, for some students at least, some of that knowledge will be remembered, and will prove important in their lives. Second, to be impressive, a piece of information does not need to be confusingly complex. In fact, a proper understanding of research findings and their methods might even help in appreciating their complexity–even if, once again, that understanding is forgotten later.

Despite these limitations, we believe that the rational impressions account offers optimism for studies of science-society interfaces, and the field of science communication in particular: Exposure to science, especially one that leaves an impression, might be the foundation of public trust in science. This means that effective science communication is essential for fostering trust in science. Low scientific literacy levels should not discourage education and communication efforts, as they are not necessarily a good indicator of the value added in terms of fostering trust in science.

Taking a broader perspective, our account fits into a picture of humans as not gullible [@mercierHowGullibleAre2017; @mercierNotBornYesterday2020]. The "failure" of the knowledge account of trust in science–the fact that science knowledge appears to not be strongly associated with trust in science–might suggest that public trust in science is, to a large extent, irrational. The notion that trust in science is irrational or easily granted may amplify concerns about the impact of misinformation: if trust lacks a solid, rational foundation, then we would expect misinformation to easily lead people astray. There is much work to be done still to understand how misinformation impacts people's beliefs, and in particular elite-driven misinformation and more subtle forms of misinformation, such as one-sided reporting. But it has been shown that people are generally able to distinguish between true and false news and, if anything, tend to be generally skeptical of news [@pfanderSpottingFalseNews2025]. As a consequence, for a better informed public, fighting for (true) information seems at least as relevant as fighting against misinformation [@acerbiResearchNoteFighting2022]. Misinformation researchers increasingly acknowledge this: A recent report on science misinformation by the National Science Foundation [@nationalacademiesofsciencesUnderstandingAddressingMisinformation2024] dedicates considerable space on developing strategies to produce better information, for example by promoting high-quality science, health, and medical journalism.

The rational impression account stresses the role of fighting for information, when it comes to fostering trust in science. Well-placed trust in science does not require profound understanding or recall of specific knowledge; but it does require exposure to good science.

# References

::: {#refs}
:::
