---
title: "Cleaning US General Social Survey (GSS) cross-section"
output: html_document
---

```{r, message=FALSE}
library(tidyverse)
library(haven) # read stata data
```

```{r, message=FALSE}
# read data
gss_cross <- read_dta("data/gss7222_r2.dta")
```

## Trust in Institutions

The GSS includes other institutional trust items that can serve as a baseline to evaluate the evolution of trust in science. 

-   [`CONFINAN`] : "Banks and financial institutions"
-   [`CONBUS`] : "Major companies"
-   [`CONCLERG`] : "Organized religion"
-   [`CONEDUC`] : "Education"
-   [`CONFED`] : "Executive branch of the federal government"
-   [`CONLABOR`] : "Organized Labor"
-   [`CONPRESS`] : "Press"
-   [`CONMEDIC`] : "Medicine"
-   [`CONTV`] : "TV"
-   [`CONJUDGE`] : "U.S. Supreme Court"
-   [`CONSCI`] : "Scientific community"
-   [`CONLEGIS`] : "Congress"
-   [`CONARMY`] : "Military"

All these variables are measured as follows: "I am going to name some institutions in this country. As far as the people running these institutions are concerned, would you say you have a great deal of confidence, only some confidence, or hardly any confidence at all in them?" {each of the 16 names of the above organizations} (1) A great deal confidence, (2) Only some confidence, (3) Hardly any confidence at all"

```{r}
# vector with all institutional trust variable names
institutional_trust_variables <- c("conbus", "conclerg", "coneduc", "conlabor", "conpress", "contv", "conmedic", "confed", "conlegis", "conjudge", "consci", "conarmy", "confinan") 

# See in which year which measure was collected
gss_cross |> 
  group_by(year) |> 
  summarize(across(all_of(institutional_trust_variables), ~ mean(.x, na.rm=TRUE)
  )
  )
```


We reverse the coding for all institutional trust variables, such that higher values represent higher trust. We also code non-valid responses as NA. 

```{r}
# Function to label values not in c(1, 2, 3) as NA
label_na_if_not_in_scale <- function(x) {
  x[!(x %in% c(1, 2, 3))] <- NA
  return(x)
}

# mini function that reverses the scale (such that higher = more confidence)
reverse_scale <- function(x) {
  return(4 - x)
}

# check
gss_cross %>% 
    mutate(across(all_of(institutional_trust_variables), 
                ~ label_na_if_not_in_scale(.x)
                )
           ) %>% 
  mutate(across(all_of(institutional_trust_variables), reverse_scale, .names = "recoded_{col}")) %>%
  select(recoded_conbus, conbus) %>%
  # make sure to select valid cases to check (in the first year,
  # trust in institutions wasn't collected)
  filter(!is.na(conbus))

# recode
gss_cross <- gss_cross %>%
  mutate(across(all_of(institutional_trust_variables), 
                ~ label_na_if_not_in_scale(.x)
  )
  ) %>% 
  mutate(across(all_of(institutional_trust_variables), reverse_scale))

table(gss_cross$consci, useNA = "always")

```

We make a long format version in which institution is a variable. 

First, we have to select only key variables (otherwise R reaches vector limit when reshaping the data frame with so many variables)

```{r}
# select variables of interest
gss_cross <- gss_cross |> 
  select(year,all_of(institutional_trust_variables))
```

We then shape the data frame to long format, with an institution variable. We also add a factor version for confidence. 

```{r}
# Create a named vector to map variable names to descriptions
institutional_trust_labels <- c(
  confinan = "Banks and financial institutions",
  conbus = "Major companies",
  conclerg = "Organized religion",
  coneduc = "Education",
  confed = "Executive branch of the federal government",
  conlabor = "Organized Labor",
  conpress = "Press",
  conmedic = "Medicine",
  contv = "TV",
  conjudge = "U.S. Supreme Court",
  consci = "Scientific community",
  conlegis = "Congress",
  conarmy = "Military"
)

# Convert variable names in dataset to labeled descriptions
gss_cross_long <- gss_cross |> 
  pivot_longer(cols = all_of(institutional_trust_variables), 
               names_to = "institution", 
               values_to = "confidence") |>
  mutate(
    institution_label = institutional_trust_labels[institution],
    confidence_factor = factor(confidence, 
                               levels = c(1, 2, 3), 
                               labels = c("Hardly any confidence at all", 
                                          "Only some confidence", 
                                          "A great deal confidence"),
                               ordered = TRUE)
  )
```


## Export

```{r}
write_csv(gss_cross_long, "data/gss_cross_cleaned.csv")
```